---
title: "Glycosylation Subpool Screen Results Summary"
date: "`r format(Sys.time(), '%Y %B %d')`"
output: 
  html_document:
    number_sections: true
    toc: true
    toc_depth: 4
    toc_float:
      collapse: TRUE
    theme: cosmo
params:
  s3bucket: "s3://[]"
  datapath: "projects/manuscript-crispr-hdt/data/subpools"
  file_screenlisting: "projects/manuscript-crispr-hdt/data/20240514-screencomparisonlisting-glycos.txt"
  outpath: "projects/manuscript-crispr-hdt/summaries/results"
  file_humanalias: "projects/manuscript-crispr-hdt/resources/aliases/gene_aliases.tsv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)

library(aws.s3)
library(aws.ec2metadata)
library(digest)
library(janitor)

library(data.table)
library(tidyr)
library(magrittr)

library(gt)
library(ggplot2)
library(cowplot)
library(pheatmap)
library(seriation)

library(fgsea)
library(msigdbr)

library(igraph)
library(visNetwork)

source("r-utils/io-utils.R")
source("r-utils/summary-utils.R")
source("r-utils/convert-utils.R")
```

```{r datapath}
s3bucket <- params$s3bucket
s3path <- file.path(s3bucket, params$datapath)
s3file <- file.path(s3bucket, params$file_screenlisting)
s3outpath <- file.path(s3bucket, params$outpath, format(Sys.time(), "%Y%m%d"), "glycosylation")

humanalias_s3file <- file.path(s3bucket, params$file_humanalias)
humanalias_dt <- format_dthumanalias(humanalias_s3file)
```

# Objective

- Summarize subpool screen results from MAGeCK files
- Note that the scoring method used here is the default MAGeCK option for calculating the null distribution (all guides).  There are only 250 non-targeting guides in the subpool.  The recommendation is at least 1000, or minimum 500 to use non-targeting guides to calculate the null distribution in the alternative MAGeCK option.  There are approximately 600 genes in the subpool.

```{r}
# read dt merge
dt_screens <- dt_froms3(s3file)

# rename pathogens
list_rename <- list("RSV-GFP5"="RSV",
										"PIV3-GFP"="PIV",
										"IAV-PR8"="IAV",
										"HRV-16"="HRV",
										"MERS-CoV-EMC2012"="MERS-CoV (1)",
										"MERS-CoV-FL2014"="MERS-CoV (2)")

catch_rename <- lapply(names(list_rename), function(virus) dt_screens[pathogen %in% virus, pathogen := list_rename[[virus]]])

dt_screens[, screen := paste0(pathogen, "_", gsub(" ", "_", cell_line))]
order_pathogen <- c("IAV",
                    "HRV",
                    "PIV",
                    "RSV",
                    "HCoV-229E",
                    "HCoV-NL63",
                    "HCoV-OC43",
                    "MERS-CoV (1)",
                    "MERS-CoV (2)",
                    "SARS-CoV-2"
                    )

dt_screens[, pathogen := factor(pathogen, levels = order_pathogen)]
dt_screens[, screen_label := paste0(pathogen, " ", cell_line, ", ", selection, " ", hpi, "HPI")]
dt_screens[, screen_label_short := paste0(pathogen, " ", cell_line)]
# purples
# https://www.color-hex.com/color-palette/28652
# oranges
# https://www.color-hex.com/color-palette/1045810
# default neg strand
dt_screens[, color_pathogen := "#e69206"]
# PIV
dt_screens[grepl("PIV", pathogen), color_pathogen := "#fd7f04"]
# HRV pos
dt_screens[grepl("HRV", pathogen), color_pathogen := "#310047"]
```

# Analysis

## Figures, Tables, and Files

### Files S5: MAGeCK Gene Summary Files

```{r}
MAGECKPATH = "mageck"
# read all mageck files
list_dtmageck <-
	lapply(1:nrow(dt_screens), function(x)
		get_genescores(report_id = dt_screens[x, report_id],
									 comparison_id = dt_screens[x, comparison_id],
									 s3path, mageckpath = MAGECKPATH))

# reorg all source mageck gene summary files in new folder with results
list_catch <- lapply(1:nrow(dt_screens), function(x)
									write_genescores(report_id = dt_screens[x, report_id],
																	 comparison_id = dt_screens[x, comparison_id],
																	 screen_name = dt_screens[x, screen],
																	 s3path, s3outpath, s3bucket, mageckpath = MAGECKPATH))

print_cols <- c("comparison_id", "screen_label")
dt_print <- dt_screens[, ..print_cols]
setnames(dt_print, "screen_label", "screen")
aws.s3::s3write_using(dt_print, 
											fwrite, 
											sep = "\t", 
											object = file.path(s3outpath, 
																				 "mageck", "screenlisting.txt"), 
											bucket = s3bucket)
```

### Volcano Plots

```{r}
# format for volcano plots
list_dtscores <-
	lapply(list_dtmageck, function(dt_mageck)
		format_mageckoutput(
			dt_mageck,
			report_cols = c(
				"comparison_id",
				"id",
				"num_guides",
				"LFC",
				"FDR",
				"pval",
				"score",
				"direction",
				"score_source"
			)
		))

list_plots <- lapply(1:nrow(dt_screens), function(i)
	plot_volcano(
		dt = list_dtscores[[i]],
		x = "LFC",
		y = "score",
		topnames = "",
		top = 20,
		x_cutoff = 0,
		y_cutoff = 0,
		xhitcol = "LFC",
		yhitcol = "FDR",
		xhit_cutoff = 0,
		yhit_cutoff = 0.05,
		ylab = "-Log10(MAGeCK Score)",
		label_fontsize = 2.5,
		main = paste0(
			dt_screens[i, pathogen],
			"|",
			dt_screens[i, cell_line],
			"\n",
			dt_screens[i, selection],
			" ",
			dt_screens[i, hpi],
			"HPI FDR<0.05"
		)
	))

files_volcano <-
	file.path(s3outpath, "volcano", dt_screens[, paste0(comparison_id, "_", screen, "_volcano.pdf")])
sapply(1:length(files_volcano), function(i)
	s3save_ggplot(
		ggp = list_plots[[i]],
		object_s3uri = files_volcano[i],
		s3bucket = s3bucket,
		width = 5,
		height = 5.2,
		units = "in",
		dpi = 300
	))
# summarize hits with different cutoffs
```

```{r}
v1 <- plot_grid(plotlist = list_plots[1:4], ncol = 2, nrow = 2)
v1
```

### Hit Summary Across Screens

```{r}
dt_scores <- rbindlist(list_dtscores)
dt_scores <-
	merge(dt_scores, dt_screens[, .(comparison_id, pathogen, screen_label)], by = "comparison_id")
```

```{r}
# use FDR < 0.05
gt_hits <-
	dt_scores[order(pathogen)][FDR < 0.05, .N, by = .(screen_label, direction)] %>% pivot_wider(
		id_cols = "screen_label",
		names_from = direction,
		values_from = N,
		values_fill = 0
	) %>% gt() |> tab_spanner(label = "Genes, FDR < 0.05", columns = c(pos, neg)) |>
	cols_label(screen_label = "CRISPRko Screen",
						 pos = "Enriched",
						 neg = "Depleted")

gt_hits
file_gthits <- "screenhits.html"
gtsave(gt_hits, file_gthits)
put_object(file_gthits, object = file.path(s3outpath, "hitsummary", file_gthits), bucket = s3bucket)
```


### PCA of pro-pathogen scores

Based on normalized gene-level ranks (guide enrichment) per screen

```{r}
# PCA on the qnorm(normalized rank) for pos
# FDR < 0.25 for any screen
# exclude genes not converted

list_dtbydirpos <- lapply(list_dtmageck, function(x) get_mageckoutputbydir(x, mydir = "pos"))
list_dtbydirneg <- lapply(list_dtmageck, function(x) get_mageckoutputbydir(x, mydir = "neg"))

dt_posneg <- rbindlist(c(list_dtbydirpos, list_dtbydirneg))[,.(id, comparison_id, FDR, pval, score, direction)]
dt_posneg <- merge(dt_posneg[,.(id, comparison_id, FDR, pval, score, direction)], dt_screens[,.(comparison_id, screen, library_species)], by = "comparison_id")
# make id less generic
setnames(dt_posneg, "id", "gene_id")

#default
human_genes <- unique(dt_posneg[library_species == "Homo sapiens", gene_id])
dt_posneg[, merge_symbol := gene_id]

# std terms based on aliases?
dt_posneg[ , human_symbol := merge_symbol]
dt_posneg[!(merge_symbol %in% human_genes), human_symbol := check_genesymbol(query_genes=ifelse(is.na(merge_symbol), gene_id, merge_symbol), dt_map=humanalias_dt)]

# focus on pos/pro-viral hits
dt_pos <- dt_posneg[direction == "pos"]
# create normalized rank for each gene in each screen (uniform distribution)
dt_pos[, norm_rank := frank(score, ties.method = "min")/.N, by = comparison_id]

# identify genes that are in all screens (unique homolog in mouse and agm)
N_CMPS = dt_pos[,length(unique(comparison_id))]
common_genes <- dt_pos[,.(n_screens = length(unique(comparison_id))), by = human_symbol][n_screens == N_CMPS, human_symbol]

dt_posconv <- dt_pos[human_symbol %in% common_genes]
# subset data FDR < 0.25
hitcommon_genes <- dt_posconv[FDR < 0.25, unique(human_symbol)]
dt_posconv <- dt_posconv[human_symbol %in% hitcommon_genes]

dt_poswide <- dt_posconv %>% pivot_wider(id_cols = human_symbol, names_from = screen, values_from = norm_rank, values_fn = min) %>% as.data.table()
df_poswide <- as.data.frame(dt_poswide)
rownames(df_poswide) <- dt_poswide[,human_symbol]
# mat for pca
mat_pos <- as.matrix(df_poswide[,c(-1)])
MAXUNIF <- max(mat_pos[mat_pos <1])
mat_pos[mat_pos == 1] = MAXUNIF
# convert to norm
mat_pos <- qnorm(mat_pos, lower.tail = F)

# pca
pca <- prcomp(t(mat_pos))

# pc var
pctvar <- 100*(pca$sdev^2)/sum(pca$sdev^2)
names(pctvar) <- colnames(pca$x)
dt_pcvar <- data.table(PC = as.numeric(gsub("PC","",names(pctvar))), var = pctvar)
p_var <- ggplot(dt_pcvar, aes(x = factor(PC), y = var)) + geom_point()
p_var <- p_var + theme_classic()
p_var <- p_var + labs(x = "PC", y = "% Variance")
p_var
s3save_ggplot(ggp = p_var, 
							object_s3uri = file.path(s3outpath, "pca", "propathogen-pca-variance.pdf"),
							s3bucket = s3bucket, width = 3, height = 5, units = "in")

# pcs
dt_pcs <- data.table(screen = rownames(pca$x))
dt_pcs[,colnames(pca$x) :=lapply(colnames(pca$x), function(i) pca$x[,i])]
dt_pcs <- merge(dt_pcs, dt_screens[,.(screen, pathogen, color_pathogen)], by = "screen")
dt_pcs[, screen := gsub("Vero_E6", "VeroE6", screen)]

# color by sense
cols_pathogen <- unique(dt_pcs[,color_pathogen])
names(cols_pathogen) <- cols_pathogen
 
p_pca <- ggplot(dt_pcs, aes(x = PC1, y = PC2, label = gsub("_","\n", screen), color = color_pathogen)) + geom_point(size = 0.5, alpha = 0.5) + ggrepel::geom_text_repel()
p_pca <- p_pca + scale_color_manual(values = cols_pathogen)
p_pca <- p_pca + theme_classic()
p_pca <- p_pca + theme(legend.position = "none")
p_pca <- p_pca + labs(x = paste0("PC1 ", round(pctvar["PC1"],1), "% Var"),
							y = paste0("PC2 ", round(pctvar["PC2"],1), "% Var"))
p_pca <-p_pca + coord_fixed(1)
p_pca

s3save_ggplot(ggp = p_pca, 
							object_s3uri = file.path(s3outpath, "pca", "propathogen-pca.pdf"),
							s3bucket = s3bucket, width = 5.5, height = 5.5, units = "in")

p_pca2 <- plot_grid(plotlist = list(p_pca, p_var), align = c("h"), rel_widths = c(1,0.5))
s3save_ggplot(ggp = p_pca2, 
							object_s3uri = file.path(s3outpath, "pca", "propathogen-pca-combined.pdf"),
							s3bucket = s3bucket, width = 8, height = 5.5, units = "in")

plot_pchm <- function(mat, genes, ...){
	
	submat <- mat[genes,]
	row_dist <- dist(submat) 
	col_dist <- dist(t(submat))
	row_order <- seriation::seriate(row_dist, method = "OLO")[[1]]$order
	col_order <- seriation::seriate(col_dist, method = "OLO")[[1]]$order
	    
	col_keys <- colnames(submat)[col_order]
	row_keys <- rownames(submat)[row_order]
	pheatmap(submat[row_order, col_order], cluster_cols = F, cluster_rows = F, ...)
}

# top 25 +/- genes in PC1

tpc1genes <- tail(sort(pca$rotation[,"PC1"]),20)
hpc1genes <- head(sort(pca$rotation[,"PC1"]),20)
topx <- c(names(tpc1genes), names(hpc1genes))

tmpfile <- paste0(tempfile(), ".pdf")
plot_pchm(mat = mat_pos, genes = names(tail(sort(abs(pca$rotation[,"PC1"])),40)), 
					filename = tmpfile, width = 4, height = 7.5)
put_object(tmpfile, object = file.path(s3outpath, "pca", "propathogen-pca-pc1heatmap-top40.pdf"), bucket = s3bucket)

tmpfile <- paste0(tempfile(), ".pdf")
plot_pchm(mat = mat_pos, genes = topx,
					filename = tmpfile, width = 4, height = 7.5)
put_object(tmpfile, object = file.path(s3outpath, "pca", "propathogen-pca-pc1heatmap-posneg-top20.pdf"), bucket = s3bucket)
```

### Figure 3a: Custom Annotion of OST Complex Genes

```{r}
genes_ost <- c( "DAD1", "DDOST", "MAGT1", "OST4", "OSTC", "RPN1", "RPN2", "STT3A", "STT3B", "TMEM258")

dt_ost <- dt_pos[human_symbol %in% genes_ost,.(human_symbol, comparison_id, FDR)]
dt_ost <- merge(dt_ost, dt_screens[, .(comparison_id, pathogen, screen_label_short)], by = "comparison_id")

aws.s3::s3write_using(dt_ost[order(FDR),.(comparison_id, screen_label_short, human_symbol, FDR)], 
											fwrite, 
											sep = "\t", 
											object = file.path(s3outpath, "subpool-crispr-ost-summaryscores.txt"), 
											bucket = s3bucket)

p <- ggplot(dt_ost, aes(x = screen_label_short, y = human_symbol, size = -log10(FDR), alpha = -log10(FDR))) + geom_point()
p <- p + theme_minimal() + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p <- p + labs(x = "", y = "")
p

s3save_ggplot(ggp = p, 
							object_s3uri = file.path(s3outpath, "Fig 3a subpool-crispr-ost-summary.pdf"),
							s3bucket = s3bucket, width = 3.5, height = 4, units = "in")

s3save_ggplot(ggp = p, 
							object_s3uri = file.path(s3outpath, "Fig 3a subpool-crispr-ost-summary.jpg"),
							s3bucket = s3bucket, device = "jpeg", width = 3.5, height = 4, units = "in", dpi = 320, bg = "white")
```


# Session Info

```{r}
sessionInfo()
```